### 2023-12-15 submission/response_to_reviewers.Rmd

---
output: pdf_document
geometry: margin=1.0in
header-includes:
 - \usepackage{setspace}
 - \doublespacing
 - \usepackage{helvet} % Helvetica font
 - \renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
 - \usepackage[T1]{fontenc}
 - \usepackage[shortcuts]{extdash}
 - \usepackage[document]{ragged2e}
---

\raggedright

I thank the editor and reviewers for their helpful comments on my manuscript. I have thoroughly addressed their comments. In addition, the revised manuscript has included alpha diversity estimates derived using the iNEXT R package, which combines non-parametric estimation and rarefaction to control for uneven effect sizes. I have also replaced the robust centered log ratio method, with the full Robust PCA method, which combines the robus CLR transformation with matrix completion. Regardless, the results are the same as robust CLR alone. When I mention line numbers in my responses to the reviewers' comments below, the line numbers indicated refer to the marked up version of the manuscript.

@@

**1. I found the result of increasing CoV in Bray-Curtis distances with increasing sequencing depth to be an interesting result given that increased sequencing depth should generally result in better confidence of what is within the underlying population. I'm wondering if the CoV results are difficult to understand due to the changing number of samples included at each sampling depth. Could this analysis be redone using only samples with sequencing depths that are above the highest rarefaction value chosen so that sample number does not vary across depth (this could be done in a subset of the samples/datasets presented). If changing sample number is not to blame are there any further speculations as to what is going on here?**

> I have added a set of panels to Figure 6 indicating the standard deviation across the samples. This hopefully makes it clear that the standard deviation between samples is decreasing, but not as fast as the mean distance between samples. On the whole, the COV values are fairly small. I have added text to this effect at L221-L224.

**2. It is clear based on the results shown that treatment groups that are highly associated with variance in sequencing depth will result in high false positive rates during diversity analysis when rarefaction is not used. However, I would argue that the simulation used does not in general reflect many natural datasets. It would be uncommon for most natural datasets to vary by treatment group by splitting sequencing depth via the median. I am wondering if this analysis could be done with varying levels of association between treatment group and sequencing depth. This will give the readers a better idea of how large this issue is in actual practice as the true association for most datasets may lay somewhere between figures 2 and 3. This would also help address how much of a concern this may be for previously published papers that do not use rarefaction during diversity analysis.**

> I appreciated this suggestion and repeated the analysis with a less severe level of confounding. In addition to the perfect confounding of treatment groups by sequencing effort, I created an additional simulation where the smallest 5% of samples were assigned to one treatment group and the largest were assigned to another; the remaining samples were randomly assigned between the groups. The new results are in Figure 3 and the previous simulation are now in Figure S4. Although the results are less severe in the new simulation, they still affirm the effects of sampling effort on the propensity to get false positive results. Additional text describing the new analysis has been inserted in the Results section at L163-L171 and the Materials & Methods section at L375-L382.

@@

**3. Based on the results of this paper it seems the key difference that is being demonstrated in these simulations as opposed to other works is the use of true rarefaction rather than subsampling. Further comments on this in the discussion would be useful to highlight why these results differ so significantly from other work such as the paper by Martino et al., in 2019 (PMID: 30801021) that introduced using robust CLR measures for microbiome data (and is quickly becoming one of the preferred choices of microbiome scientists). Furthermore, it would be nice to see whether the use of true rarefaction on the datasets presented in Martino et al., 2019 resolves the conflicting views between this report and Martino et al's findings on robust CLR measures. Overall, a clarification on how these two publications come to differing conclusions about best practices when comparing beta diversity would be helpful to a large audience within the microbiome field.**

> I appreciate the reviewer's suggestion to look closer at the Martino et al study and their RPCA method. In the revised manuscript I used RPCA rather than the simple RCLR transformation. In the process of my work, I found a number of problems with the design and analysis of the benchmarking experiments performed in their study. While I mention the differences between my analysis and the Martino et al. and McMurdie & Holmes studies in the revised manuscript (L345-L355), a more comprehensive analysis is underway and is beyond the scope of the current study.

@@

**4. I am wondering whether any of the differences in the results derived from studies such as Martino et al., 2019 (PMID: 30801021), can be attributed to sequence processing pipeline. While the author makes convincing arguments that removal of low abundance sequences may not always be appropriate, I am wondering how it would impact the results shown. Especially given the widespread use of this practice in the current literature as indicated by the author in the discussion. This would be especially important given the strong language around this practice in the discussion without primary data to back it up within this manuscript.**

> My responses to the previous points partially address this question. The data processing pipeline do not factor in for interpreting the results of this analysis since all of my benchmarking analyses used the same data for a dataset and each dataset was processed with the same pipeline. The methods don't know what type of processing was done upstream. Part of the value of using 12 different datasets was to provide diversity in community shapes. I could have used purely synthetic distributions as others have done (e.g. Martino et al., McMurdie & Holmes) and obtained similar results.

**5. On line 164 the report indicates finding differences in richness between samples that should have the same OTU counts given the way the enrichment sampling was done (increasing sequencing counts not raw number of OTUS in the sample). It would make sense that metrics that consider abundances to change but not richness. Comments on this would be appreciated. Furthermore, this section jumps around between two different models of enrichment (abundance enrichment vs. raw count enrichment). A clearer separation of these two analyses would be appreciated to make parsing the text easier as in some cases it is unclear which enrichment strategy was used for the result that is being talked about.**

> The text in this paragraph has been edited to make it more clear what simulation is being used for each analysis. There is also text in the Discussion highlighting the value of using the Shannon or inverse Simpson's diversity indices over richness because they depend less on rare taxa (see paragraph starting at L255).

@@

**Minor comments:**

**1. Well, the current consensus is that sequence depth is mostly attributed to noise (Gloor et al., 2017. PMID: 29187837). It would be nice to comment on some recent papers that indicate that through certain methods of library preparation sequencing depth may be correlated with sample biomass and how this reflects with the current thoughts in this manuscript (Cruz et al., 2021. PMID: 33717032, Munich et al., 2022. PMID: 36396943).**

> I have added the citations the reviewer suggested to the text at L355-L340 and fleshed out the discussion in this paragraph.

@@

**2. It would be nice if confidence intervals could be included in the correlation plots in Figure 1. Furthermore, if understood correctly Figures 2-3 y-axis should be the mean false positive rate over 100 dataset simulations? If so a measure of data spread would be appropriate for these values.**

> The caption for Figure 1 indicates that the size of the confidence interval was smaller than the plotting symbol. As stated in the captions for Figures 2 and 3, "The false positive rate was the number of times a dataset yielded a significant result." Therefore, it does not have a confidence interval.

@@

**3. While I agree that the rarefaction value chosen is not arbitrary and is usually a balance between keeping samples and having the highest sequencing depth possible (as indicated in the discussion of this manuscript). It would be nice to give stronger guidelines around choosing an appropriate value to highlight more robust ways of choosing rarefaction depths. One thought that came across my mind would be the use of CoV calculations from repeated subsampling (rarefaction) to determine appropriate sequencing depths. Or perhaps just a simple calculation of good's coverage at various sequence depths (similar to rarefaction plots used now)?**

> I'm reluctant to add more than what I already have at L308-L315. In that text, I mention using Good's coverage values as a check to make sure that the level of coverage is acceptable. In the revised manuscript I did add alpha diversity metrics that incorporate sampling depth and coverage to extrapolate (iNEXT) (L316-L327). Although these methods perform better than their "raw" counterparts, there are still effects of sampling effort.

@@

**4. Based on these results and various studies on microbiome biases, alpha diversity metrics such as richness do not indicate the true richness of a sample due to a variety of issues included sequencing depth. A comment on this in the discussion would be helpful, for those that are less seasoned in the microbiome field.**

> I have added a sentence to this effect in the Discussion at L274-L277.

@@

**This study addresses a controversy that has been playing out in the literature regarding the processing of 16S rRNA gene amplicon data. Some researchers contend that rarefaction (the random subsampling with replacement to account for differences in sequencing depth among samples) is undesireable because it "wastes data". Schloss does a comprehensive analysis of how rarefaction performs compared to other ways of processing the samples (including using raw counts). He compares both alpha and beta diversity metrics using both a mock dataset and a collection of published datasets from a range of ecosystems. In nearly all cases, rarefaction**

**My only "major" critique relates to the core purpose of the article: to convince readers that rarefaction is a better way to process data than those proposed by other researchers. After reading the article multiple times, I still don't yet comprehend why the other researchers came to their conclusions. At some level, they must have convinced themselves that the use of raw read count (or other methods) out-perform rarefaction. Did they really not do the same comparisons that are done here? Without reading those carefully, I can't tell. Schloss does mention that they removed rare OTUs in their analysis. Is this the main reason why they did not detect the superior performance of rarefaction? I think just a brief explanation for Schloss' interpretation of why his conclusions are different from McMurdle/Holmes would be even more convincing for readers who don't want to dig too deeply into other papers. There is some discussion in the introduction, but I'd like to see it in the actual discussion section, after the full presentation of the new analysis.**

**Kudos. Love the running title. Tell us how you really feel!**

> My pleasure :). I have added text to the closing paragraph of the Discussion (L345-L355) to elaborate on why the results of my benchmarking studies differ so strikingly with those that have preceded mine.

**Minor comments**

**Line 69 - This implies that McMurdle/Holmes used an incorrect interpretation of differences between rarefying and rarefaction. This is an opportunity to clarify! Do you agree with their interpretation (but want to promote the use of rarefaction specifically)?**

@@

**Line 72 - Do you mean that the reanalysis did not remove 15% of the samples?**

> I have clarified this paragraph of the Introduction (L64-L78). My paper describing the re-analysis of WNWN has recently been accepted for publication at *mSphere* and goes into far greater details regarding the problems with the original study.

@@

**Line 72 - Change "method" to "methods"**

> This has been corrected

**Line 78 - Did they use the term "toy" in their work? This comes off as unnecessarily snarky. Would it be appropriate to use "simulated" instead?**

> No snark was intended. "Toy" is jargon in the field for a simplistic example desined to make a point. To avoid confusion I have used "overly simplistic" instead.

### 2023-12-15 submission/manuscript.Rmd
Editing / submitted

---
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
csl: asm.csl
geometry: margin=1.0in
header-includes:
 - \usepackage{upgreek}
 - \usepackage{booktabs}
 - \usepackage{longtable}
 - \usepackage{graphicx}
 - \usepackage{array}
 - \usepackage{multirow}
 - \usepackage{wrapfig}
 - \usepackage{float}
 - \usepackage{colortbl}
 - \usepackage{pdflscape}
 - \usepackage{tabu}
 - \usepackage{threeparttable}
 - \usepackage{threeparttablex}
 - \usepackage[normalem]{ulem}
 - \usepackage{makecell}
 - \usepackage{setspace}
 - \doublespacing
 - \usepackage[left]{lineno}
 - \linenumbers
 - \modulolinenumbers
 - \usepackage{helvet}
 - \renewcommand*\familydefault{\sfdefault}
 - \usepackage[T1]{fontenc}
 - \usepackage[shortcuts]{extdash}
 - \usepackage[document]{ragged2e}
---

@@

The ability to generate millions of 16S rRNA gene amplicon and metagenomic sequence reads has allowed researchers to multiplex multiple samples on the same sequencing run by pooling separate PCRs that can be deconvoluted later based on index (aka barcode) sequences that are embedded into the sequence of the PCR primers [@Sogin2006; @Caporaso2010; @Kozich2013]. Unfortunately, it is common to observe variation in the number of sequence reads from each sample vary by as much as 100-fold (e.g, see Figure S1). This occurs because pooling of DNA from multiple PCRs is fraught with numerous opportunities for technical errors to compound leading to a skewed distribution. Aside from developing better methods of pooling DNA, the question of how to control for uneven sequencing effort in microbial ecology studies has become controversial.

@@

In 2014, McMurdie and Holmes [@McMurdie2014] announced that "rarefying" of microbial community data was "statistically inadmissible" because it omits valid data. In their simulations, they observed that rarefying reduced the statistical power to correctly cluster samples into the same treatment groups based on beta diversity metrics. Although the detail was lost on many subsequent researchers, they did not describe rarefaction, but rarefying [@Schloss2023wnwn]. According to McMurdie and Holmes, rarefying is a single subsampling of an OTU abundance table whereas rarefaction repeats the subsampling step many times. Furthermore, their simulations penalized the rarefied data by removing 15% of the samples and evaluated clustering quality using a clustering algorithm that performed worse for rarefied data. According to my reanalysis of rarefaction using the original simulation framework, rarefaction outperformed other normalization method being used in 2014 for both alpha and beta diversity metrics [@Schloss2023wnwn]. Others have also critiqued the original work [@Weiss2017]. Regardless, there is a general sense in the community that rarefying and rarefaction should not be used with microbiome data.

Since McMurdie and Holmes published their simulations, others have developed alternative approaches to control for uneven sequencing effort in amplicon sequencing studies. For alpha diversity metrics, Willis used toy datasets to demonstrate that one could estimate the richness for each sample in a dataset and use those values for statistical comparisons [@Willis2019]. Non-parametric estimators of richness and diversity [@Chao2003; @Chao2016] and parametric estimators of richness [@Willis2015] have been used in microbial ecology studies. For beta diversity metrics at least four approaches have been pursued. First, one could use relative abundance values where the observed number of sequences in an OTU is divided by the total number of sequences in the sample [@Lin2020]. Second, normalization strategies have been developed where the relative abundance is multiplied by the size of the minimum desired sequencing effort and fractional values are reapportioned to the OTUs to obtain integer values [@Beule2020; @Paulson2013]. Third, a variety of center log-ratio methods have been developed where the compositional nature of the OTU counts is removed and used to calculated Euclidean distances (aka Aitchison distances) [@Quinn2019; @Paulson2013; @Lin2020; @Martino2019; @Costea2014]. This strategy is purported to control for uneven sequencing effort [@Martino2019; @Quinn2018]; however, some have noticed that this feature breaks down under certain conditions [@teBeest2021]. Finally, as mentioned above, variance stabilization transformations have been recommended to generate values that can be used to calculate Euclidean distances [@McMurdie2014].

The ongoing controversy over the use of rarefaction and the recent development of alternative strategies to control for uneven sequencing effort caused me to question how these methods compared to each other using a simulation framework that overcame the issues with the McMurdie and Holmes study. My analysis included 16S rRNA gene sequence data from from 12 studies that characterized the variation in bacterial communities from diverse environments (Table 1 and Figure S1). The sequences were assigned to OTUs using a standard pipeline and their frequencies and the number of sequences found in each sample were used to generate simulated communities and treatment effects. For each dataset and simulation, 100 replicate datasets were generated and used as input to each of the strategies for controlling for uneven sequencing effort. My overall conclusion was that rarefaction outperformed the alternative strategies.

@@

***Without rarefaction, metrics of alpha diversity are sensitive to sequencing effort.*** To test the sensitivity of various approaches of measuring alpha diversity to sequencing effort, I generated null models using OTU counts and sequencing depths from 12 studies. Under a null model, each community from the same dataset would be expected to have the same alpha diversity regardless of the sequencing effort. I measured the richness of the communities in each dataset without any correction, using scaled ranked subsampling (SRS) normalized OTU counts, with estimates based on non-parametric and parametric approaches, and using rarefaction (e.g. Figure S2). For each dataset, all of the approaches, except for rarefaction, showed a strong correlation between richness and the number of sequences in the sample (Figure 1A). Next, I assessed diversity using the Shannon diversity index and the inverse Simpson diversity index without any correction, using normalized OTU counts, and rarefaction; I also used a non-parametric estimator of Shannon diversity. The correlation between sequencing depth and the diversity metric was not as strong as it was for richness and the inverse Simpson diversity values were less sensitive than the Shannon diversity values; however, the correlation to the diversity metrics calculated with rarefaction were the lowest for all of the metrics and studies (Figure 1A). The alpha-diversity metrics calculated with rarefaction consistently demonstrated a lack of sensitivity to sequencing depth.

***Without rarefaction, metrics of beta diversity are sensitive to sequencing effort.*** To test the sensitivity of various approaches of measuring beta diversity to sequencing effort, I used the same null models used for studying the sensitivity of alpha diversity. Under a null model, the ecological distance between any pair of samples would be the same regardless of the difference in the number of sequences observed in each sample (e.g., Figure S3). First, I calculated the Jaccard distance coefficient between all pairs of communities within a dataset. The Jaccard distance coefficient is the fraction of OTUs that are unique to either community and does not account for the abundance of the OTUs. Jaccard distances were calculated using the uncorrected OTU counts, with rarefaction, relative abundances, and following normalization using cumulative sum scaling (CSS) and SRS. Only the distances calculated with rarefaction showed a lack of sensitivity to sequencing effort (Figure 1B). Second, I analyzed the sensitivity of the Bray-Curtis distance coefficient, which is a popular metric that incorporates the abundance of each OTU. Similar to what I observed with the Jaccard coefficient, only the data normalized with rarefaction showed a lack of sensitivity to sequencing effort (Figure 1B). Third, I calculated the Euclidean distance on raw OTU counts where the central log-ratio (CLR) was calculated (i.e., Aitchison distances) using matrix completion to replace zero counts (Robust CLR), adding a pseudo-count of one to all OTU counts prior to calculating the CLR (One CLR), adding a pseudo-count of one divided by the total number of sequences obtained for the community (Nudge CLR), and imputing the value of zero counts (Zero CLR). The Aitchison distances were all strongly sensitive to sequencing effort (Figure 1B). Finally, I used the variance-stabilization transformation (VST) from DESeq2 prior to calculating Euclidean distances. Again, there was a strong sensitivity to sequencing effort (Figure 1B). Although Euclidean distances are not typically used on raw or rarefaction-normalized count data in ecology [@Legendre2012], Euclidean distances calculated with rarefaction were not sensitive to sequencing effort. Across each of the beta diversity metrics and approaches used to account for uneven sequencing effort and sparsity, rarefaction was the least sensitive approach to differences in sequencing effort.

***Rarefaction limits the detection of false positives when sequencing effort and treatment group are confounded.*** Next, I investigated the impact of the various strategies and metrics on falsely detecting a significant difference using the the same communities generated from the null model in the analysis of alpha and beta diversity metrics. To test for differences in alpha and beta diversity I used the non-parametric Wilcoxon test and non-parametric permutation-based multivariate analysis of variance (PERMANOVA). First, I employed an unbiased null treatment model to measure the false detection rate, which should not have meaningfully differed from 5%. Indeed, for each dataset and alpha and beta diversity metric and strategy for accounting for uneven sequencing, approximately 5% of the tests yielded a significant result (Figure 2). Second, I employed a biased null treatment model where the treatment group was completely confounded with the number of sequences in each sample (Figure S4). Under this model, only the data normalized with rarefaction consistently resulted in a 5% false positive rate for alpha and beta diversity metrics. I then used a less severe bias where the smallest 5% of samples were assigned to one treatment group and the largest 5% were assigned to another; the other samples were randomly assigned to both treatment groups. Although the sensitivity of the false positive rate to the level of confounding varied across methods of quantifying alpha and beta diversity, rarefaction was the only method to consistently control the false positive rate (Figure 3). These results aligned with the observed sensitivity of alpha and beta diversity metrics to sequencing effort and underscore the value of rarefaction.

***Rarefaction preserves the statistical power to detect differences between treatment groups.*** To assess the impact of different approaches to control for uneven sequencing effort I performed two additional simulations. In the first simulation, I implemented a skewed abundance distribution model to create two treatment groups for each dataset that were each populated with half of the samples each with the same number of sequences as the samples had in the observed data. The two treatment groups varied in their structure such that one had the same abundances as the null distribution above and the other had 10% of its OTUs randomly selected to increase their counts by 5%. The power to detect differences in richness between the two simulated treatment groups by all approaches was low (Figure 4A). This was likely because the approach for generating the perturbed community did not necessarily change the number of OTUs in each treatment group. Regardless, the simulations testing differences in richness using the Rice and Stream datasets had the greatest power when the richness data were calculated with rarefaction. To explore this further, a richness-adjusted community model was created by removing 3% of the OTUs from a null model model. As suggested by the first simulation, the richness data  calculated with rarefaction had a higher statistical power than the other approaches when measuring richness (Figure 5). The simulations testing the power to detect differences in Shannon diversity also showed that rarefaction performed better than the other methods (Figure 4A). When testing for differences in the Inverse Simpson diversity index the the difference between rarefaction and the other methods was negligible (Figure 4A). For tests of beta diversity I found that rarefaction was the most reliable approach to maintain statistical power to detect differences between two communities. Among the tests using the Jaccard and Bray-Curtis metrics, raw count data and CSS normalized data had little power relative to using rarefaction, relative abundance, and SRS to normalize the uneven sequencing depths. The differences in power between counts normalized with rarefaction, relative abundance, and SRS data was small, but if there were differences, the power obtained using rarefaction was greater than the other methods. Among the tests using Euclidean distances, using raw counts and CLR and DESeq2 transformed data had little power relative to the distances calculated using rarefaction and relative abundances. This power-based analysis of the simulated communities using different methods of handling uneven sample sizes demonstrated the value of rarefaction for preserving the statistical power to detect differences between treatment groups for measures of alpha and beta diversity.

@@

extrapolate <- cor_line %>%
  mutate(ninety = abs(mean - 0.90),
         ninetyfive = abs(mean - 0.95),
         ninetynine = abs(mean - 0.99)) %>%
  group_by(dataset) %>%
  summarize(ninety = nseqs[which.min(ninety)],
            ninetyfive = nseqs[which.min(ninetyfive)],
            ninetynine = nseqs[which.min(ninetynine)]) %>%
  mutate(ninetyfive = ninetyfive / ninety - 1,
         ninetynine = ninetynine / ninety - 1) %>%
  summarize(m = round(mean(ninetyfive), 1),
            n = round(mean(ninetynine), 1)
            )
```

>>>>

---
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
  word_document: default
csl: asm.csl
geometry: margin=1.0in
header-includes:
 - \usepackage{upgreek}
 - \usepackage{booktabs}
 - \usepackage{longtable}
 - \usepackage{graphicx}
 - \usepackage{array}
 - \usepackage{multirow}
 - \usepackage{wrapfig}
 - \usepackage{float}
 - \usepackage{colortbl}
 - \usepackage{pdflscape}
 - \usepackage{tabu}
 - \usepackage{threeparttable}
 - \usepackage{threeparttablex}
 - \usepackage[normalem]{ulem}
 - \usepackage{makecell}
 - \usepackage{setspace}
 - \doublespacing
 - \usepackage[left]{lineno}
 - \linenumbers
 - \modulolinenumbers
 - \usepackage{helvet}
 - \renewcommand*\familydefault{\sfdefault}
 - \usepackage[T1]{fontenc}
 - \usepackage[shortcuts]{extdash}
 - \usepackage[document]{ragged2e}
---

@@

The ability to generate millions of 16S rRNA gene amplicon and metagenomic sequence reads has allowed researchers to multiplex multiple samples on the same sequencing run by pooling separate PCRs that can be deconvoluted later based on index (aka barcode) sequences that are embedded into the sequence of the PCR primers [@Sogin2006; @Caporaso2010; @Kozich2013]. Unfortunately, it is common to observe variation in the number of sequence reads from each sample vary by as much as 100-fold (e.g, see Figure S1). This occurs because pooling of DNA from multiple PCRs is fraught with numerous opportunities for random sampling errors to compound leading to a skewed distribution. Aside from developing better methods of pooling DNA, the question of how to control for uneven sequencing effort in microbial ecology studies has become controversial.

@@

In 2014, McMurdie and Holmes [@McMurdie2014] announced that "rarefying" of microbial community data was "statistically inadmissible" because it omits valid data. In their simulations, they observed that rarefying reduced the statistical power to correctly cluster samples into the same treatment groups based on beta diversity metrics. I have recently shown that their analysis had a number of important problems [@Schloss2023wnwn]. First, although the detail was lost on many subsequent researchers, they did not describe rarefaction. Instead they described rarefying as a single subsampling of an OTU abundance table whereas rarefaction repeats the subsampling step many times. Furthermore, their simulations penalized the rarefied data by removing 15% of the samples but did not remove those samples when considernig their other methods. They also evaluated the accuracy of clustering samples using a clustering algorithm that performed worse for rarefied data. According to my reanalysis of rarefaction using their original simulation framework, but using all of the samples, rarefaction outperformed the other normalization methods they described for both alpha and beta diversity metrics [@Schloss2023wnwn]. Others have also critiqued the original work [@Weiss2017].

Since McMurdie and Holmes published their simulations, others have developed alternative approaches to control for uneven sequencing effort in amplicon sequencing studies. For alpha diversity metrics, Willis used overly simplistic datasets to demonstrate that one could estimate the richness for each sample in a dataset and use those values for statistical comparisons [@Willis2019]. Non-parametric estimators of the minimum richness in a community and diversity [@Chao2003; @Chao2016; @Chao2014] and parametric estimators of richness [@Willis2015] have been used in microbial ecology studies. One non-parametric approach, iNEXT, combines rarefaction on samples more than a threshold number of sequences and extrapoliation on samples below the threshold [@Chao2014]. This method has become popular in traditional ecology studies, but has not been widely used amongst microbial ecologists. For beta diversity metrics at least four approaches have been pursued as alternatives to rarefaction. First, one could use relative abundance values where the observed number of sequences in an OTU is divided by the total number of sequences in the sample [@Lin2020]. Second, normalization strategies have been developed where the relative abundance is multiplied by the size of the minimum desired sequencing effort and fractional values are reapportioned to the OTUs to obtain integer values [@Beule2020; @Paulson2013]. Third, a variety of center log-ratio methods have been developed where the compositional nature of the OTU counts is removed and used to calculated Euclidean distances (aka Aitchison distances) [@Quinn2019; @Paulson2013; @Lin2020; @Martino2019; @Costea2014]. This strategy is purported to control for uneven sequencing effort [@Martino2019; @Quinn2018]; however, some have noticed that this feature breaks down under certain conditions [@teBeest2021]. Finally variance stabilization transformations have been recommended to generate values that can be used to calculate Euclidean distances [@McMurdie2014].

The ongoing controversy over the use of rarefaction and the recent development of alternative strategies to control for uneven sequencing effort caused me to question how these methods compared to each other using a simulation framework that overcame the issues with the McMurdie and Holmes study [@McMurdie2014]. My analysis included 16S rRNA gene sequence data from 12 diverse environments (Table 1 and Figure S1). The sequences were assigned to OTUs using a standard pipeline and their frequencies and the number of sequences found in each sample were used to generate simulated communities and treatment effects. For each dataset and simulation, 100 replicate datasets were generated and used as input to each of the strategies for controlling for uneven sequencing effort. My overall conclusion was that rarefaction outperformed the alternative strategies.

@@

***Without rarefaction, metrics of alpha diversity are sensitive to sequencing effort.*** To test the sensitivity of various approaches of measuring alpha diversity to sequencing effort, I generated null models using OTU counts and sequencing depths from 12 studies. Under a null model, each community from the same dataset would be expected to have the same alpha diversity regardless of the sequencing effort. I measured the richness of the communities in each dataset without any correction, using scaled ranked subsampling (SRS) normalized OTU counts, with estimates based on non-parametric and parametric approaches, and using rarefaction (e.g. Figure S2). For each dataset, all of the approaches, except for rarefaction and possibly the size-based iNEXT estimates, showed a strong correlation between richness and the number of sequences in the sample (Figure 1A). The size-based iNEXT estimates were correlated with sampling effort for the Marine and Soil datasts, which had fewer samples and more uniform sequencing depth distributions than the other datasets. Next, I assessed diversity using the Shannon diversity index and the inverse Simpson diversity index without any correction, using normalized OTU counts, non-parametric estimation, and rarefaction. The correlation between sequencing depth and the Shannon diversity metric was not as strong as it was for richness. The inverse Simpson diversity values was not sensitive to sampling effort (Figure 1A). The alpha-diversity metrics calculated with rarefaction consistently demonstrated a lack of sensitivity to sequencing depth.

***Without rarefaction, metrics of beta diversity are sensitive to sequencing effort.*** To test the sensitivity of various approaches of measuring beta diversity to sequencing effort, I used the same null models used for studying the sensitivity of alpha diversity. Under a null model, the ecological distance between any pair of samples would be the same regardless of the difference in the number of sequences observed in each sample (e.g., Figure S3). First, I calculated the Jaccard distance coefficient between all pairs of communities within a dataset. The Jaccard distance coefficient is the fraction of OTUs that are unique to either community and does not account for the abundance of the OTUs. Jaccard distances were calculated using the uncorrected OTU counts, with rarefaction, relative abundances, and following normalization using cumulative sum scaling (CSS) and SRS. Only the distances calculated with rarefaction showed a lack of sensitivity to sequencing effort (Figure 1B). Second, I analyzed the sensitivity of the Bray-Curtis distance coefficient, which is a popular metric that incorporates the abundance of each OTU. Similar to what I observed with the Jaccard coefficient, only the data normalized with rarefaction showed a lack of sensitivity to sequencing effort (Figure 1B). Third, I calculated the Euclidean distance on raw OTU counts where the central log-ratio (CLR) was calculated (i.e., Aitchison distances) by adding a pseudo-count of one to all OTU counts prior to calculating the CLR (One CLR), adding a pseudo-count of one divided by the total number of sequences obtained for the community (Nudge CLR), imputing the value of zero counts (Zero CLR), and using CLR on non-zero counts followed by matrix completion on zero counts (Robust Principle Components Analysis [PCA]). The Aitchison distances were all strongly sensitive to sequencing effort (Figure 1B). Finally, I used the variance-stabilization transformation (VST) from DESeq2 prior to calculating Euclidean distances. Again, there was a strong sensitivity to sequencing effort (Figure 1B). Although use of Euclidean distances is not advised with raw or rarefaction-normalized count data in ecology [@Legendre2012], Euclidean distances calculated with rarefaction were not sensitive to sequencing effort. Across each of the beta diversity metrics and approaches used to account for uneven sequencing effort and sparsity, rarefaction was the least sensitive approach to differences in sequencing effort.

***Rarefaction limits the detection of false positives when sequencing effort and treatment group are confounded.*** Next, I investigated the impact of the various strategies and metrics on falsely detecting a significant difference using the the same communities generated from the null model in the analysis of alpha and beta diversity metrics. To test for differences in alpha and beta diversity I used the non-parametric Wilcoxon test and non-parametric permutation-based multivariate analysis of variance (PERMANOVA). First, I employed an unbiased null treatment model to measure the false detection rate, which should not have meaningfully differed from 5%. Indeed, for each dataset and alpha and beta diversity metric and strategy for accounting for uneven sequencing, approximately 5% of the tests yielded a significant result (Figure 2). Second, I employed a biased null treatment model where the treatment group was confounded with the number of sequences in each sample such that the 50% of samples with the least sequencing effort were in one treatment group and the others were in another treatment group (Figure S4). Under this model, only the data normalized with rarefaction consistently resulted in a 5% false positive rate for alpha and beta diversity metrics. I then used a less severe bias where the smallest 5% of samples were assigned to one treatment group and the largest 5% were assigned to another; the other samples were randomly assigned to both treatment groups. Although the sensitivity of the false positive rate to the level of confounding varied across methods of quantifying alpha and beta diversity, rarefaction and the iNEXT size-based estimates for alpha diveristy were the only methods to consistently control the false positive rate (Figure 3). These results aligned with the observed sensitivity of alpha and beta diversity metrics to sequencing effort and underscore the value of rarefaction.

***Rarefaction preserves the statistical power to detect differences between treatment groups.*** To assess the impact of different approaches to control for uneven sequencing effort I performed two additional simulations. In the first simulation, I implemented a skewed abundance distribution model to create two treatment groups for each dataset that were each populated with half of the samples each with the same number of sequences as the samples had in the observed data. The two treatment groups varied in their structure such that one had the same abundances as the null distribution above and the other had 10% of its OTUs randomly selected to increase their counts by 5%. The power to detect differences in richness between the two simulated treatment groups by all approaches was low (Figure 4A). This was likely because the approach for generating the perturbed community did not necessarily change the number of OTUs in each treatment group; the differences primarily affected the distribution of the sequences across OTUs. Regardless, the simulations testing differences in richness using the Rice and Stream datasets had the greatest power when the richness data were calculated with rarefaction. To explore this further, a richness-adjusted community model was created by randomly selecting and removing 3% of the OTUs from a null model. As suggested by the first simulation, the richness data calculated with rarefaction consistently had the highest statistical power among the approaches used to measure richness followed by the iNEXT size-based estimates (Figure 5). Returning to simulations where treatment groups varied in their structure, the power to detect differences in Shannon diversity was the lowest when using raw relative abundances than when using the other methods (Figures 4A and 5). When testing for differences in the Inverse Simpson diversity index the the difference between rarefaction and the other methods was negligible (Figures 4A and 5). For tests of beta diversity, I found that rarefaction was the most reliable approach to maintain statistical power to detect differences between two communities (Figure 4B). Among the tests using the Jaccard and Bray-Curtis metrics, raw count data and CSS normalized data had little power relative to using rarefaction, relative abundance, and SRS to normalize the uneven sequencing depths. The differences in power between counts normalized with rarefaction, relative abundance, and SRS data was small, but if there were differences, the power obtained using rarefaction was greater than the other methods. Among the tests using Euclidean distances, using raw counts and CLR and DESeq2 transformed data had little power relative to the distances calculated using rarefaction and relative abundances. This power-based analysis of the simulated communities using different methods of handling uneven sample sizes demonstrated the value of rarefaction for preserving the statistical power to detect differences between treatment groups for measures of alpha and beta diversity.

@@

extrapolate <- cor_line %>%
  mutate(ninety = abs(mean - 0.90),
         ninetyfive = abs(mean - 0.95),
         ninetynine = abs(mean - 0.99)) %>%
  group_by(dataset) %>%
  summarize(ninety = nseqs[which.min(ninety)],
            ninetyfive = nseqs[which.min(ninetyfive)],
            ninetynine = nseqs[which.min(ninetynine)]) %>%
  mutate(ninetyfive = ninetyfive / ninety - 1,
         ninetynine = ninetynine / ninety - 1) %>%
  summarize(m = round(mean(ninetyfive), 0),
            n = round(mean(ninetynine), 0)
            )
```

***Increased rarefaction depth reduces intra-sample variation in alpha and beta diversity.*** One concern with using rarefaction is the perceived loss of sequencing information when a large fraction of data appears to be removed when the community with the greatest sequencing depth is sampled to the size of the community with the least (e.g., the smallest sample in the `r loss$nice_name` dataset had `r loss$min_max`% of sequences that were in the largest sample). To assess the sensitivity of alpha and beta diversity metrics to rarefaction depth, I again used the dataset generated using the null models, but used rarefaction with each community to varying sequencing depths (Figure 6). The richness values increased with sequencing effort as rare OTUs would continue be detected. In contrast, the Shannon diversity and Bray-Curtis values plateaued with increased sequencing effort. This result was expected since increased sequencing would lead to increased precision in the measured relative abundance of the OTUs. Next, I measured the coefficient of variation (i.e., the mean divided by the standard deviation) between samples for richness, Shannon diversity, and Bray-Curtis distances. Although the mean richness appeared to increase unbounded with sequencing effort, the coefficient of variation for each dataset was relatively stable. In general, the coefficient of variation increased slightly with sequencing depth only to decline once smaller samples were removed from the analysis at higher sequencing depths. Interestingly, the coefficient of variation between Shannon diversity values decreased towards zero with increased sequencing effort and the coefficient of variation between Bray-Curtis distances tended to increase slightly for samples where the standard deviation did not decrease as quickly as the mean distance between samples. Regardless, the coefficients of variation were relatively small. This analysis indicates that there are benefits to increased sequencing depths.
